What should I notice? Evaluating memorability of events to generate surprising causal candidates.

### 2021-08-16 paper.tex
some changes

The rest of this paper is organized as follows: we first briefly introduce some notions of
Algorithmic Information Theory and its applications in section~\ref{sec:theory},
then we explicit our methods of computing the unexpectedness of past events in
cyber-physical systems~\ref{sec:computing}. We illustrate this approach by
providing an implementation relying on a few predicates on events from a smart
home simulation~(section \ref{sec:example}). In section~\ref{sec:related} we briefly
review other related works using complexity theory or trying to explain smart
homes and how our work can be linked with them before exploring potential
extensions of our work in section \ref{sec:future}

@@

It is important to note that the notion of Kolmogorov complexity is not related
to computation complexity, as there is no requirement on the execution time
of the programs, only their length in bits matters for the computation of
complexity. In fact, it can be shown that Kolmogorov complexity is not
computable\cite{li_introduction_2008}. The sketch of the proof is the following:
since the definition of complexity from eq. \ref{eq:kolm_def} requires to find
the shortest of all programs on a given Turing machine outputting $s$, it
requires to run all programs up to a given length and compare their result to
$s$. However, being able to do so means that we would be able to tell whether
these programs terminate. Since the termination of programs is famously
undecidable, by extension, the computation of Kolmogorov complexity is
impossible. One can however easily approximate it with upper bounds, by
exhibiting a program outputting $s$.

@@

This bridge between Algorithmic Information Theory and human perception of
complexity is extended to define the notion of simplicity and unexpectedness,
which are considered of uttermost importance in cognitive science\cite{chater_simplicity_2003}.
\cite{dessalles2011coincidences} proposes a formal definition of the unexpectedness $U(e)$
of an event, as the difference between an a-priori expected causal
complexity $C_{w}(e)$ and the actual observed complexity $C(e)$.
\begin{equation}
  \label{eq:unexpected}
  U(e) = |C_{w}(e) - C(e)|
\end{equation}
This result comes from the understanding that, while Kolmogorov complexity is
computed with Turing machine, it can be used as an elegant proxy for modeling
information processing in the human brain, and thus helps design a notion of
simplicity or complexity of events.

@@

The definition used for description complexity in eq. \ref{eq:desc_k} allows for
a direct implementation, which is shown in alg. \ref{alg:complex_iter}. This
algorithm operates iteratively: starting with the base memory $\mathcal{M}$
(line 1), we apply all possible predicate concepts $\pi$ from a given set
$\mathcal{P}$ and programs $k$ (lines 6-7), up to a given length
$\mathtt{max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We
then store the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array
$\mathtt{future_{explore}}$. At the end of the iteration, the results of the
filters become the memories which will be explored during the next iteration
(lines 21--23). Each pass thus explore retrieval paths of increasing length.
When a singleton memory is reached, the complexity of its unique element is
upper-bounded with the length of the corresponding retrieval path (line 14).

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{./figures/roc}
  \caption{Experimental ROC curve (True Positive Rate against False Positive Rate) for a
    classifier based on our memorability score.}
  \label{fig:roc}
\end{figure}
\subsubsection{Abduction}

@@

\section{Related Works}
\label{sec:related}
\textbf{TODO}

>>>>

The rest of this paper is organized as follows: we first briefly introduce
some notions of Algorithmic Information Theory and its applications in
section~\ref{sec:theory}, then we explicit our methods of computing the
unexpectedness of past events in cyber-physical systems~\ref
{sec:computing}. We illustrate this approach by providing an implementation
relying on a few predicates on events from a smart home simulation~
(section \ref{sec:example}). In section~\ref{sec:related} we briefly review
other related works using complexity theory or trying to explain smart homes
and how our work can be linked with them before exploring potential
extensions of our work in section \ref{sec:future}

@@

It is important to note that the notion of Kolmogorov complexity is not related
to computation complexity, as there is no requirement on the execution time of
the programs, only their length in bits matters for the computation of
complexity. In fact, it can be shown that Kolmogorov complexity is not
computable\cite{li_introduction_2008}. The sketch of the proof is the
following: since the definition of complexity from eq. \ref
{eq:kolm_def} requires to find the shortest of all programs on a given Turing
machine outputting $s$, it requires to run all programs up to a given length
and compare their result to $s$. However, being able to do so means that we
would be able to tell whether these programs terminate. Since the termination
of programs is famously undecidable, by extension, the computation of
Kolmogorov complexity is impossible. One can however easily approximate it with
upper bounds, by exhibiting a program outputting $s$.

@@

This bridge between Algorithmic Information Theory and human perception of
complexity is extended to define the notion of simplicity and unexpectedness,
which are considered of uttermost importance in cognitive science\cite
{chater_simplicity_2003}.
\cite{dessalles2011coincidences} proposes a formal definition of the
 unexpectedness $U(e)$ of an event, as the difference between an a-priori
 expected causal complexity $C_{w}(e)$ and the actual observed complexity $C
 (e)$.
\begin{equation}
  \label{eq:unexpected}
  U(e) = |C_{w}(e) - C(e)|
\end{equation}
This result comes from the understanding that, while Kolmogorov complexity is
computed with Turing machine, it can be used as an elegant proxy for modeling
information processing in the human brain, and thus helps design a notion of
simplicity or complexity of events.

@@

The definition used for description complexity in eq. \ref{eq:desc_k} allows for
a direct implementation, which is shown in alg. \ref{alg:complex_iter}. This
algorithm operates iteratively: starting with the base memory $\mathcal{M}$
(line 1), we apply all possible predicate concepts $\pi$ from a given set
$\mathcal{P}$ and programs $k$ (lines 6-7), up to a given length $\mathtt
{max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
{explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explore retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{./figures/roc}
  \caption{Experimental ROC curve (True Positive Rate against False Positive
   Rate) for a classifier based on our memorability score. Measures consider 23
   manually flagged events as memorable (events added to the background noise
   as described in Section~\ref{sec:example}.)}
  \label{fig:roc}
\end{figure}
\subsubsection{Abduction}

@@

\section{Related Works}
\label{sec:related}
Our work is intended to be integrated into larger-scale frameworks to monitor 
and detect events in complex environment such as smart homes. In these works, the
approach to smart homes is often regarded as self-organizing systems \cite{kramer_rigorous_2009,kounev_notion_2017}. As such, they present capacities of adaptation to new goals, 
new components, new environment. A commonly used approach is the principle of 
autonomic system, which minimizes user's intervention for management of the 
system \cite{kounev_notion_2017,kephart_vision_2003}.

We want to inscribe our work among other classical concurrent approaches to 
abduction. In situations where more data is available, we could for instance rely
on correlation or causal inference from known relations \cite{peters_elements_2017,fadiga_or_2021}.
Previous relations between inference and complexity have been studied. In fact, the 
case of inference was one of the motivations for R. Solomonoff to introduce his 
universal algorithmic probability \cite{solomonoff_formal_1964} as a tool to 
reach an idealized inference machine, creating the notion of complexity concomitantly 
to Kolmogorov. Subsequently, notions of complexity re-emerged in causal inference: 
\cite{janzing_causal_2010} found, when a causal link exists between two random variables,
considering the direct joint probability is simpler, in terms of Kolmogorov complexity, 
than the inverse direction.

The topic of finding events from streams of time series data has already been explored in many ways. 
\cite{aggarwal_outlier_2017} provides good review of modern approaches and techniques
in the field.

While all these works advocate in favor of a strong link between complexity and the 
discovery of causes, not other works extends the notion up to the point we propose in this
work, using complexity only as a tool for inference.
